---
title: "Human Development Index"
author: "David Jackson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(scales)
```

## Human Development Index

```{r}
human <- read_csv("./human-development-index.csv") %>%
  clean_names()
```
```{r}
summary(human$human_development_index)
```
```{r}


# Select countries to visualize
countries <- c("Norway", "United States", "India", "China")

# Filter data for selected countries
hdi_countries <- human %>%
  filter(entity %in% countries)

# Create line plot
ggplot(hdi_countries, aes(x = year, y = human_development_index, color = entity)) +
  geom_line() +
  scale_x_continuous(breaks = seq(min(human$year), max(human$year), by = 10)) +
  labs(title = "Changes in HDI over Time", x = "Year", y = "HDI Score", color = "Country") +
  theme_bw()
```
#### 2.	Compare HDI scores across countries: 

Create a bar chart or box plot that shows the distribution of HDI scores across countries in a selected year. This can help you compare the levels of human development across different countries and regions.

```{r}

# Select year for comparison
Year <- 2019

# Filter data for selected year
hdi_year <- human %>%
  filter(Year == year)

# Create box plot
ggplot(hdi_year, aes(x = entity, y = human_development_index)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = paste("Distribution of HDI Scores Across Countries in", Year),
       x = "Country", y = "HDI Score") +
  theme_bw()

```
```{r}
ggplot(hdi_year, aes(x = entity, y = human_development_index)) +
  geom_boxplot() + coord_flip()
```

Here's some R code that you can use to create a ranking plot and table to analyze trends in HDI rankings over time:

```{r}

# Pivot data to create ranking table
ranking_table <- human %>%
  select(entity, year, human_development_index) %>%
  pivot_wider(names_from = year, values_from = human_development_index) %>%
  mutate(rank_1990 = rank(`1990`, ties.method = "min"),
         rank_2000 = rank(`2000`, ties.method = "min"),
         rank_2010 = rank(`2010`, ties.method = "min"),
         rank_2020 = rank(`2020`, ties.method = "min"))
```
```{r}
# Create ranking plot
ggplot(ranking_table, aes(rank_1990, rank_2020, color = entity)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  scale_x_reverse() +
  scale_y_reverse() +
  theme_minimal() +
  labs(title = "HDI Ranking Changes from 1990 to 2020",
       x = "Ranking in 1990",
       y = "Ranking in 2020",
       color = "Country")

# View ranking table
ranking_table

```


